---
title: "Getting Started with CyteTypeR"
author: "Badran Elshenawy"
date: today
format:
  html:
    theme: darkly
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    embed-resources: true
    self-contained: true
    title-block-banner: true
engine: knitr
execute:
  eval: false
  warning: false
  message: false
---

## Introduction

This guide demonstrates **CyteTypeR**, an API-driven tool for automated cell type annotation in scRNA-seq data. CyteTypeR prepares cluster markers, aggregated metadata, and UMAP embeddings, then submits them to the CyteType API for annotation.

-   **What CyteTypeR returns**:
    -   Updated **Seurat object** with predicted cell type annotations in metadata
    -   **Interactive HTML report** with cluster-level visualizations and marker analysis
-   **CyteTypeR GitHub**: <https://github.com/NygenAnalytics/CyteTypeR>
-   **Python version**: <https://github.com/NygenAnalytics/CyteType>
-   **Example report**: <https://nygen-labs-prod--cytetype-api.modal.run/report/34fac9e9-3c43-4c46-95f4-6b2994e57ada?cluster=7&tab=markers>

## Installation

CyteTypeR is installed from GitHub via `devtools`. All other dependencies are available from CRAN or Bioconductor.

```{r}
#| label: installation

devtools::install_github("NygenAnalytics/CyteTypeR")
```

## Data Import

-   **Dataset**: IFNB stimulation dataset from `SeuratData`
    -   Contains PBMCs from control and IFN-beta stimulated conditions
    -   Commonly used as a benchmark dataset for scRNA-seq integration and annotation workflows
-   **Loading method**: `SeuratData::LoadData()` returns a Seurat object ready for processing

```{r}
#| label: data-import

# Libraries ----------

library(Seurat)
library(SeuratData)
library(SCpubr)
library(tidyseurat)
library(CyteTypeR)
library(glmGamPoi)
library(tictoc)

# Load data ----------

ifnb <- SeuratData::LoadData(ds = "ifnb")
```

## Normalization and Dimensionality Reduction

-   **Normalization**: `SCTransform()` with `vst.flavor = "v2"`
    -   Replaces the traditional `NormalizeData` + `FindVariableFeatures` + `ScaleData` pipeline
    -   Uses regularized negative binomial regression to model technical noise
    -   `vst.flavor = "v2"` uses the improved glmGamPoi-based model for better variance stabilization
-   **PCA**: Reduces high-dimensional gene expression space to principal components capturing major biological variation
-   **UMAP**: Projects PCA embeddings into 2D for visualization
    -   `dims = 1:20` uses the first 20 PCs, balancing signal capture against noise inclusion

```{r}
#| label: normalization-dimreduction

# Libraries ----------

library(Seurat)
library(glmGamPoi)

# Inputs ----------

# ifnb object from previous chunk

# Normalization ----------

ifnb <- SCTransform(ifnb, vst.flavor = "v2")

# Dimensionality reduction ----------

ifnb <- RunPCA(ifnb)
ifnb <- RunUMAP(ifnb, dims = 1:20)
```

| Parameter | Value | Justification |
|-----------------------|------------------|-------------------------------|
| `vst.flavor` | `"v2"` | Improved variance stabilization via glmGamPoi backend |
| `dims` (UMAP) | `1:20` | Standard range; captures major biological variation while excluding noisy higher PCs |

## Clustering

-   **Graph construction**: `FindNeighbors()` builds a shared nearest-neighbor (SNN) graph from PCA space
    -   Uses the same `dims = 1:20` as UMAP for consistency
-   **Clustering algorithm**: `FindClusters()` applies the Louvain algorithm on the SNN graph
    -   `resolution = 0.4` produces moderate granularity — lower values yield fewer, broader clusters

```{r}
#| label: clustering

# Libraries ----------

library(Seurat)

# Inputs ----------

# ifnb object from previous chunk

# Graph construction ----------

ifnb <- FindNeighbors(ifnb, dims = 1:20)

# Clustering ----------

ifnb <- FindClusters(ifnb, resolution = 0.4)
```

| Parameter | Value | Justification |
|-----------------------|------------------|-------------------------------|
| `dims` | `1:20` | Matches dimensionality used for UMAP |
| `resolution` | `0.4` | Moderate granularity; avoids over-splitting clusters in this dataset |

## Differential Expression

-   **Method**: `FindAllMarkers()` performs pairwise differential expression testing for each cluster against all other cells
    -   Identifies marker genes that distinguish each cluster from the rest
    -   `group.by = "seurat_clusters"` ensures markers are computed per Seurat cluster identity
-   **Purpose**: Marker genes are the primary input for CyteTypeR's cell type prediction

```{r}
#| label: differential-expression

# Libraries ----------

library(Seurat)

# Inputs ----------

# ifnb object from previous chunk

# Find cluster markers ----------

clust_markers <- FindAllMarkers(ifnb, group.by = "seurat_clusters")
```

## Prepare Data for CyteTypeR

-   **`PrepareCyteTypeR()`** transforms the Seurat object and markers into the JSON structure expected by the CyteType API:
    -   **Cluster labels**: Mapping of cell indices to cluster identities
    -   **Aggregated metadata**: Per-cluster percentage breakdowns of each categorical metadata column (e.g., condition, sample)
    -   **UMAP coordinates**: 2D embeddings for all cells
    -   **Top marker genes**: Top N genes per cluster ranked by differential expression significance
-   **Report metadata**: A named list providing context for the interactive HTML report

```{r}
#| label: prepare-cytetype

# Libraries ----------

library(CyteTypeR)

# Inputs ----------

# ifnb object and clust_markers from previous chunks

# Prepare API payload ----------

prepped_data <- PrepareCyteTypeR(
  ifnb,
  clust_markers,
  n_top_genes = 10,
  group_key = "seurat_clusters",
  aggregate_metadata = TRUE,
  coordinates_key = "umap"
)

# Report metadata ----------

metadata <- list(
  title = "My scRNA-seq analysis of the ifnb dataset",
  run_label = "initial_analysis",
  experiment_name = "ifnb_human_samples_study"
)
```

| Parameter | Value | Justification |
|-----------------------|------------------|-------------------------------|
| `n_top_genes` | `10` | Top 10 markers per cluster balances specificity with noise reduction |
| `group_key` | `"seurat_clusters"` | Uses the Seurat-assigned cluster identities |
| `aggregate_metadata` | `TRUE` | Computes per-cluster percentage breakdowns of metadata columns |
| `coordinates_key` | `"umap"` | Sends UMAP embeddings for the interactive report visualization |

## Automated Cell Type Annotation

-   **`CyteTypeR()`** submits the prepared data to the CyteType API
    -   `study_context` provides biological context that improves annotation accuracy (tissue type, experiment description)
    -   `metadata` populates the interactive HTML report header
-   **Returns**:
    -   Seurat object with predicted cell type labels added to metadata columns
    -   URL to an interactive HTML report with per-cluster annotation details and marker visualizations

```{r}
#| label: annotation

# Libraries ----------

library(CyteTypeR)
library(tictoc)

# Inputs ----------

# ifnb object, prepped_data, and metadata from previous chunks

# Submit to CyteTypeR API ----------

tic("CyteTypeR Run")
ifnb_annotated <- CyteTypeR(
  obj = ifnb,
  prepped_data = prepped_data,
  study_context = "ifnb dataset from SeuratData package",
  metadata = metadata
)
toc()
```

| Parameter | Value | Justification |
|-----------------------|------------------|-------------------------------|
| `study_context` | Dataset description | Provides biological context to improve annotation accuracy |
| `metadata` | Named list with title, run_label, experiment_name | Populates the interactive report header |

## Summary

CyteTypeR provides a streamlined workflow for automated cell type annotation:

-   **API-driven**: Offloads annotation to a specialized service rather than relying on local reference datasets
-   **Seurat-native**: Accepts and returns Seurat objects with no format conversion required
-   **Report generation**: Produces interactive HTML reports alongside programmatic annotations
-   **Minimal setup**: Requires only standard Seurat preprocessing (normalization, clustering, marker detection) as input